Implementing XBF: An Efficient Self-Describing Binary Format

### 2023-08-15 docs/paper.tex
Update paper

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xurl}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{microtype}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

On the web, data is largely interchanged in one of two formats, XML (Extensible Markup Language)\cite{xml_spec} or JSON (JavaScript Object Notation)\cite{json_spec}. These formats have two key features that have led to their popularity, human-readability and being self-describing. These features have inherent inefficiencies, which XBF (Xtensible Binary Format)\cite{xbf_spec} intends to solve through carefully chosen compromises.

@@

A self-describing format is a format in which a recipient is not required to have a schema for the incoming data, and the names and types of the data being sent can be determined from the data itself. The following is an example of XML:

@@

A human-readable format is one where data is encoded entirely as ASCII or UTF-8 characters, and formatted in such a way that a person could easily read or write it without the need of a machine. Human-readability brings with it two inefficiencies, the need to parse text data into a binary format that a computer understands, and additional overhead from whitespace characters.

@@

Similar to CSV (cite here?), we send a single header up front, followed by all the data. For a primitive type there isn't a large savings here, but there are massive savings with vectors and structs. With a vector the metadata is a discriminant indicating a vector is being sent, which is then followed by the metadata for its internal type. This process continues recursively if the inner type (or any of its inner types) contains a vector or struct, ensuring that all metadata necessary to represent the data coming afterward is received up front. The data is then sent in a single continuous block, something after this...

@@

The reference implementation for XBF\cite{xbf_impl} is written in Rust\cite{rust}. Rust was chosen for the following reasons:
\begin{IEEEitemize}
	\item Rust is statically typed and compiled, and allows for similarly efficient code generation to C/C++.
	\item A powerful type system with tagged unions built-in\cite{enums} such that dynamic dispatch and virtual method tables weren't required, and structural pattern matching was used instead.
	\item Rust provides statically checked memory safety through the concept of ownership\cite{ownership}.
	\item A built-in testing system in the standard library\cite{testing}.
	\item Errors are represented as values rather than throwing exceptions\cite{results}.
\end{IEEEitemize}

@@

@@

@@

@@

In Rust there is an important distinction between an owned String, and a string slice\cite{strings}. A string slice is similar to the previously mentioned slices, except it consists of valid UTF-8 code points. It is a view into a block of memory, and its size cannot be modified. If growable string is required, that's where the String type is used. It makes the same UTF-8 guarantee as a string slice, but has the same properties as a Vec where it keeps track of its capacity, will reallocate if its capacity is exceeded, similar to a dynamic array. As with the decision to use a slice in XbfStruct, a string slice is utilized to save on memory usage, as the metadata for a struct shouldn't need to change, thereby making a capacity field irrelevant.

@@

XBF performed significantly faster than all except one of the other formats tested, despite minimal effort spent optimizing the reference implementation's runtime performance. Additionally, XBF has the smallest overhead of the formats tested, indicating that it would make a good choice for IOT (internet of things) systems which need to transfer data in a low bandwidth scenario. Based on the overall

@@

In the self-describing binary format space, there are still many improvements that could be made in the future to XBF. Sizes for variable length

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{microtype} 	% fixes word distribution in citations
\usepackage{xurl} 			% allows citation urls to wrap around lines
\usepackage{bytefield} 	% for shoing the memory layout of types 
\usepackage{xpatch} 		% fixes bytefield looking wonky with IEEE formatting

\xpretocmd{\bytefield}{\setlength{\lineskip}{1pt}}{}{}

% weird IEEE stuff included in the template, not questioning it
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

@@

On the web, data is largely interchanged in one of two formats, XML (Extensible Markup Language)\cite{xml_spec} or JSON (JavaScript Object Notation)\cite{json_spec}. These formats have two key features that have led to their popularity, human-readability and being self-describing. These features have inherent inefficiencies, which XBF (Xtensible Binary Format) intends to solve through carefully chosen compromises.

@@

A self-describing format contains metadata and does not require a schema for incoming data, and the names and types of the data being sent can be determined from the data itself. The following is an example of XML:

@@

A human-readable format is one where data is encoded entirely as ASCII or UTF-8 characters, and formatted in such a way that a person could easily read or write it without the need of a machine. Human-readability brings with it two inefficiencies, the need to parse text data into a binary format that a computer understands, and additional overhead from the inherent amount of space it takes to encode data in ASCII. For example, a 32-bit number is stored in 4 bytes natively, but a variable amount of space when in ASCII. If the number being stored is 1, then ASCII only takes a single byte (only a single digit). Once more than 4 digits is required (numbers over 9999), ASCII will always be less efficient than a natively encoded number. In a worst case scenario, 32-bit numbers can store a number up to around 4.2 billion, which would take 10 bytes to store in plain text.

\iffalse
	\begin{table}[htbp]
		\caption{Plain Text vs Binary}\label{binary_vs_text}
		\begin{center}
			\begin{tabular}{|c|c|c|}
				\hline
				\textbf{Value} & \textbf{Plain Text Bytes} & \textbf{Binary Bytes} \\
				1              & 1                         & 4                     \\
				9999           & 4                         & 4                     \\
				10000          & 5                         & 4                     \\
				\hline
			\end{tabular}
		\end{center}
	\end{table}
\fi

@@

\subsection{Primitive Types}

There are a total of 17  primitive types:

\begin{IEEEitemize}
	\item Boolean
	\item U8, U16, U32, U64, U128, U256
	\item I8, I16, I32, I64, I128, I256
	\item F32, F64
	\item Bytes
	\item String (UTF-8)
\end{IEEEitemize}

All integers should be sent in little endian format, with the least significant bit first. For signed integers, they should be represented in two's complement format. These two formats are chosen because they are the most common way integers are interacted with in modern x86 and ARM processors, and should minimize the amount of conversion required when sending and receiving data. Similarly, floating point numbers should be sent as 32 or 64 bit IEEE 754 floating point numbers.

Strings should be sent as a sequence of bytes that correspond to UTF-8 code points. They should first send their length as an unsigned 64-bit integer (in little endian format), followed by the corresponding number of bytes contained within the string. Bytes have the same specification as strings, but with the exception that they do not have to be a valid sequence of UTF-8 code points. Lengths were chosen to be 64-bits to simplify the format, but at the cost of efficiency when sending many small sequences.

\subsection{Primitive Metadata}

Primitive metadata should be sent as a single byte discriminant value. After receiving one of these discriminant values, the client should read the following byte(s) and interpret them as the type given by the discriminant.

\begin{table}[htbp]
	\caption{Primitive Metadata Discriminants}\label{discriminants}
	\begin{center}
		\begin{tabular}{|c|c|}
			\hline
			\textbf{Type} & \textbf{Discriminant} \\
			\hline
			Boolean       & 0                     \\
			U8            & 1                     \\
			U16           & 2                     \\
			U32           & 3                     \\
			U64           & 4                     \\
			U128          & 5                     \\
			U256          & 6                     \\
			I8            & 7                     \\
			I16           & 8                     \\
			I32           & 9                     \\
			I64           & 10                    \\
			I128          & 11                    \\
			I256          & 12                    \\
			F32           & 13                    \\
			F64           & 14                    \\
			Bytes         & 15                    \\
			String        & 16                    \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

Strings should always be the final value in the list. The value given to strings
is used by Vectors and Structs to determine what their discriminant value should
be.

\subsection{Primitive Examples}

In the following examples, each type will first be shown with its discriminant value followed by its actual value, with the byte representation underneath them.

\subsubsection{16-Bit Unsigned Integer}

\begin{center}
	\begin{bytefield}{24}
		\bitbox{8}{Disc} & \bitbox{16}{1024} \\
		\bitbox{8}{2} & \bitboxes{8}{04} \\
	\end{bytefield}
\end{center}

\subsubsection{String}

\newcommand{\rot}[1]{\rotatebox{90}{#1}}

\begin{center}
	\begin{bytefield}[bitheight=\widthof{"Disc"}]{28}
		\bitbox{2}{\rot{Disc}} & \bitbox{16}{length} & \bitboxes{2}{hello} \\
		\bitboxes{2}{{16} {2} {0} {0} {0} {0} {0} {0} {0} {\rot{104}} {\rot{101}} {\rot{108}} {\rot{108}} {\rot{111}}}
	\end{bytefield}
\end{center}

\subsection{Vector Type}

Vectors are a homogenous list of values that has a known, variable length. Vectors should first include their length as an unsigned 64-bit integer (the same as Strings and Bytes), followed by the corresponding number of elements. The type contained within a Vector is not sent to the client. That information is carried in the metadata.

\subsection{Vector Metadata}

Similarly to primitives, a single byte discriminant value signifying a Vector is incoming is sent. This discriminant value should be 1 greater than that of the discriminant value for Strings. Following this, metadata information for the internal type contained within the Vector will be sent. This process may continue recursively with nested types of Vectors and Structs. The length of a vector or the data contained within the vector must not be sent.

\subsection{Vector Example}

\textit{Vector of 16-Bit Unsigned Integers}

DV = Discriminant of a Vector

D16 = Discriminant of a 16-Bit Unsigned Integer

\begin{center}
	\begin{bytefield}{26}
		\bitbox{2}{\rot{DV}} & \bitbox{2}{\rot{D16}} & \bitbox{16}{length} & \bitboxes{4}{{42} {1024}} \\
		\bitboxes{2}{{17} {2} {2} {0} {0} {0} {0} {0} {0} {0} {42} {0} {0} {4}}
	\end{bytefield}
\end{center}

\subsection{Struct Type}

A Struct is an aggregate type containing a name as well as named fields. A struct may not contain duplicate field names. Should a Struct be sent like this anyway, it should be considered malformed and not be constructed on the receiving end. Fields of a Struct are sent in sequence in the order they are listed in the Struct's metadata. When a struct is serialized it should not send any sort of name information (such as its name or field names), how many fields it has, nor should it send any type information about its fields. That information is carried in the metadata.

\subsection{Struct Metadata}

A discriminant value should first be sent, similarly to primitives (following the same size requirement). This discriminant value should be 1 greater than that of the discriminant value for Vectors.

Following this, the name of the Struct should be sent, using the same format as primitive strings are sent (unsigned 64-bit length and then the bytes). Next, send the number of fields contained within the Struct as a U64, the same as all other lengths. Finally, the fields of the Struct should be sent, first the name of the field as a String, then immediately after the metadata for the type of the field. This process may continue recursively with nested types of Structs or Vectors. These name and type pairs will be sent until there are no more fields left in the Struct.

\subsection{Struct Example}

\begin{lstlisting}[language=C]
struct MyStruct {
	x: i8,
	y: i8
}

let my_struct = MyStruct {
	x: 42,
	y: -5,
};
\end{lstlisting}

@@

The reference implementation for XBF\cite{xbf_impl} is written in Rust\cite{rust}. Rust was chosen for the following reasons:

\begin{IEEEitemize}
	\item Rust is statically typed and compiled, and allows for similarly efficient code generation to C/C++.
	\item A powerful type system with tagged unions built-in\cite{enums} such that dynamic dispatch and virtual method tables weren't required, and structural pattern matching was used instead.
	\item Rust provides statically checked memory safety through the concept of ownership\cite{ownership}.
	\item A built-in testing system in the standard library\cite{testing}.
	\item Errors are represented as values rather than throwing exceptions\cite{results}.
\end{IEEEitemize}

@@

\begin{table}[htbp]
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			\multicolumn{3}{|c|}{XbfType}     \\
			\hline
			XbfPrimitive & XbfVec & XbfStruct \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

\begin{table}[htbp]
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			\multicolumn{3}{|c|}{XbfMetadata}                         \\
			\hline
			XbfPrimitiveMetadata & XbfVecMetadata & XbfStructMetadata \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

@@

\begin{center}
	\begin{bytefield}{32}
		\wordbox{1}{XbfVec} \\
		\bitbox{8}{Metadata} & \bitbox{24}{Vec\textless XbfType\textgreater} \\
		\bitbox{8}{Rc} & \bitbox{8}{*XbfType} & \bitbox{8}{Length} & \bitbox{8}{Capacity}
	\end{bytefield}
\end{center}

@@

\begin{center}
	\begin{bytefield}{32}
		\wordbox{1}{XbfStruct} \\
		\bitbox{8}{Metadata} & \bitbox{24}{Box\textless [XbfType]\textgreater} \\
		\bitbox{8}{Rc} & \bitbox{12}{*XbfType} & \bitbox{12}{Length}
	\end{bytefield}
\end{center}

@@

In Rust there is an important distinction between an owned String, and a string slice\cite{strings}. A string slice is similar to the previously mentioned slices, except it consists of valid UTF-8 code points. It is a view into a block of memory, and its size cannot be modified. If growable string is required, that's where the String type is used. It makes the same UTF-8 guarantee as a string slice, but has the same properties as a Vec where it keeps track of its capacity, and will reallocate if its capacity is exceeded, similar to a dynamic array. As with the decision to use a slice in XbfStruct, a string slice is utilized to save on memory usage, as the metadata for a struct shouldn't need to change, thereby making a capacity field irrelevant.

\begin{center}
	\begin{bytefield}{24}
		\wordbox{1}{String} \\
		\bitbox{8}{*u8} & \bitbox{8}{Length} & \bitbox{8}{Capacity} \\
	\end{bytefield}

	\begin{bytefield}{16}
		\wordbox{1}{\&str (string slice)} \\
		\bitbox{8}{*u8} & \bitbox{8}{Length}
	\end{bytefield}
\end{center}

@@

The first result to address is the difference between the number of bytes in the original CSV file and the number of bytes received by the client when asking for CSV data. Upon examination, it was found that the original CSV always contained 8 total digits of precision, whereas when it was serialized only the number of digits required to represent a number was sent. As an example, in the original data the number 83.5 would be represented as 83.500000, but when serialized would omit the trailing zeroes, as they were unnecessary.

XBF performed significantly faster than all except one of the other formats tested, despite minimal effort spent optimizing the reference implementation's runtime performance. This can likely be attributed working entirely with binary data, which avoids needing to do complex and expensive conversions between plain text and native binary formats, as well as requires sending less data in general. TODO: more here

Additionally, XBF has the smallest overhead of the formats tested, indicating that it would make a good choice for IOT (internet of things) systems which need to transfer data in a low bandwidth scenario. Based on the overall TODO: more here

@@
